---
title: "Undocumented immigrants contribute $100 billion through taxes"
author: "Diego Perdomo for Futuro Media Group"
output: html_document
date: "2025-06-07"
editor_options: 
  markdown: 
    wrap: 72
---

In May, federal courts permitted immigration agencies to access
confidential, personal information of up to seven million undocumented
immigrant taxpayers. This is the first instance where the Internal
Revenue Service has shared confidential information for the purposes of
immigration investigation, according to the National Immigration Forum.

Required by law to file taxes regardless of citizenship, America's
undocumented immigrants, estimated at 11 million in 2022, are
responsible for nearly \$100 billion of contributions to federal, state
and local governments in 2022, according to a July report from Institute
on Taxation and Economic Policy (ITEP).

<iframe title="NY undocumented pay highest rate to local, state taxes" aria-label="Map" id="datawrapper-chart-ADPu9" src="https://datawrapper.dwcdn.net/ADPu9/5/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="556" data-external="1">

</iframe>

```{=html}
<script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>
```

Since 1996, the IRS has permitted people with noncitizen status to file
their taxes through the Individual Taxpayer Identification Number. The
identification number, which permitted undocumented immigrants to file
their taxes and receive select tax credits, fell under scrutiny in
recent federal court hearings and prompted the landmark collaboration
between the Department of Homeland Security and Department of Treasury.

With new questions of undocumented immigrants' relationship to taxes and
rising confidentiality concerns, I think it would be timely and relevant
to Futuro’s audience to investigate:

-   The portfolio of state and local taxes undocumented immigrants
    contribute to with \$37.3 billion

    -   Data visualizations like charts and maps to describe state-level
        rates for each tax revenue type (income, federal, sales, etc.)

-   How undocumented taxpayer contributions differ in makeup from
    citizen taxpayers

    -   The portfolio of taxes undocumented immigrants contribute
        compared to the U.S. total of tax revenue

Given the missing data on the exact counts of undocumented immigrants,
it would be interesting to examine:

-   How other tax policy think tanks measure immigrant contributions

    -   Particularly the methodology of these studies and what
        assumptions they make from the holes in their data.

-   Other instances of immigrant contribution to federal, state and
    local revenue

    -   Commerce and census data pertaining to income and nativity

Through this research, I intend to add new analysis to data that has
been discussed among spaces that report on immigrants and taxes for
almost 11 months and prepare information for followup articles as the
story develops.

#Preparation \## Load software libraries

```{r setup, include=FALSE}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)
library(DT)
```

## Scraping html for tables from ITEP website

From \$96.7 billion, a third is spent on state and local taxes. In their
"Tax Payments by Undocumented Immigrants" report, ITEP created an
estimate of the undocumented immigrant population through microdata
provided by the census's five-year recurring American Community Survey.
With that data, they made further estimates to approximate their tax
contributions to federal, state and local governments. To dissect this
data further, we decided to scrape it from the website's table and clean
it for state by state comparisons. The charts, Appendix Table 1
(state/local tax totals) and Appendix Table 5 (population estimates),
were read and copied from the site using an XPath scraping method.

### XPath for tax table

```{r}
#State local tax contribs from undoc immigrants in 2022
# Read the HTML file from the URL
url <- "https://itep.org/undocumented-immigrants-taxes-2024/"
html <- read_html(url)

# Extract the relevant information using XPath
undoc_imm_taxes <- html %>%
  html_nodes(xpath = "/html/body/div/div/div[2]/div/div[1]/div[1]/div[4]/table") %>%
  html_nodes("td") %>%
  html_text() %>%
  matrix(ncol = 7, byrow = TRUE) %>%
  as.data.frame(stringsAsFactors = FALSE)

# Assign column names
new_headers <- as.character(undoc_imm_taxes[1, ])
undoc_imm_taxes_clean <- undoc_imm_taxes[-1, ] # Remove the first row
colnames(undoc_imm_taxes_clean) <- new_headers
```

### XPath for population table

```{r}
#Undocumented immigrants population and income data, 2022
#Read html for other table
undoc_imm_income_pop <- html |> 
  html_nodes(xpath = "/html/body/div[1]/div/div[2]/div/div[1]/div[1]/div[8]/table") |> 
  html_nodes("td") |> 
  html_text() |> 
  matrix(ncol = 3, byrow = TRUE) |> 
  as.data.frame(stringsAsFactors = FALSE)

#Column Names
undoc_imm_income_pop_clean <- undoc_imm_income_pop 

colnames(undoc_imm_income_pop_clean) <- undoc_imm_income_pop[1,]

undoc_imm_income_pop_clean <- undoc_imm_income_pop_clean[-1,]
rownames(undoc_imm_income_pop_clean) <- 1:nrow(undoc_imm_income_pop_clean)
```

# Cleaning

```{r, warning=FALSE, message=FALSE}
#Cleaning state local tax contribs from undocumented immigrants, 2022
undoc_imm_taxes <- undoc_imm_taxes_clean |> 
  clean_names() |> 
  #remove dollar signs
  mutate(
    sales_and_excise_taxes=as.numeric(parse_number(sales_and_excise_taxes)),
    property_taxes=as.numeric(parse_number(property_taxes)),
    personal_and_business_income_taxes=as.numeric(parse_number(personal_and_business_income_taxes)),
    other_taxes=as.numeric(parse_number(other_taxes)),
    total_state_and_local_taxes=as.numeric(parse_number(total_state_and_local_taxes)),
     effective_tax_rate=as.numeric(parse_number(effective_tax_rate)),
    )
write.csv(undoc_imm_taxes, "../data/state_local_tax_undoc_2022.csv")

#Remember to credit the source in your articles: Institute on Taxation and Economic Policy
```

```{r}
#Cleaning undocumented immigrant population and income data
undoc_imm_income_pop <- undoc_imm_income_pop_clean |> 
  clean_names() |> 
  #remove dollar signs
  mutate(
    population=as.numeric(parse_number(population)),
    aggregate_income=as.numeric(parse_number(aggregate_income))
    )

write_csv(undoc_imm_income_pop, "../data/aggregate_income_undoc_2022.csv")

#Remember to credit the source in your articles: Institute on Taxation and Economic Policy
```

# Explore the data types

```{r}

glimpse(undoc_imm_taxes)

```

# Code for later Fix dates

```{r}
# data <- data |> 
#   mutate(date = lubridate::mdy(contribution_date))
```

# Data exploration: taxes paid

## State and Local Taxes

Upon using analysis to sort the data by each column descending, I
realized that many of the tax totals were just population

```{r}
# Income total by State
undoc_imm_taxes |> 
  select(state, personal_and_business_income_taxes) |> 
  arrange(desc(personal_and_business_income_taxes)) |> 
  head(5)
# Property total by State
undoc_imm_taxes |> 
  select(state, property_taxes) |> 
  arrange(desc(property_taxes)) |> 
  head(5)
# Sales total by State
undoc_imm_taxes |> 
  select(state, sales_and_excise_taxes) |> 
  arrange(desc(sales_and_excise_taxes)) |> 
  head(5)
# Other total by State
undoc_imm_taxes |> 
  select(state, other_taxes) |> 
  arrange(desc(other_taxes)) |> 
  head(5)
# State/local total by State
undoc_imm_taxes |> 
  select(state, total_state_and_local_taxes) |> 
  arrange(desc(total_state_and_local_taxes)) |> 
  head(5)
```

```{r, echo=FALSE}
#Per capita
undoc_imm_taxes_income <- inner_join(undoc_imm_taxes, undoc_imm_income_pop, by = "state")

#Sales - likely dependent on sales tax rates and population
undoc_imm_taxes_income <- undoc_imm_taxes_income |> 
  mutate(sales_capita = sales_and_excise_taxes/population,
         property_capita = property_taxes/population,
         personal_and_business_income_capita = personal_and_business_income_taxes/population,
         other_capita = other_taxes/population,
         total_capita = total_state_and_local_taxes/population) 

dt_undoc_tax_income <- undoc_imm_taxes_income |> 
  select(state,personal_and_business_income_capita,sales_capita, property_capita, total_capita) |> 
  rename(State="state",
         `Individual and Business Income Rate` = "personal_and_business_income_capita",
         `Sales Rate` = "sales_capita",
         `Property Rate` = "property_capita",
         `Total Rate` = "total_capita")

#Export and do visual analysis on datawrapper to visualize trends
write.csv(undoc_imm_taxes_income, "../data/taxes_capita_undoc_2022.csv")

datatable(dt_undoc_tax_income,
          caption = htmltools::tags$caption(
            style = 'caption-side: top; text-align: center; font-size: 24px; font-weight: bold; padding: 10px;',
            "Contributions to local and state taxes from undocumented immigrants"
          ),
          options = list(
            pageLength = 20,
            order = list(list(5, 'desc'))  # Sort by 4th column (Total Rate) descending
          )) |>  
  formatCurrency(c('Individual and Business Income Rate','Sales Rate','Property Rate','Total Rate'),currency = "$", interval = 3, mark = ",")
```

# Questions

Write down questions you want answered:

-   What is the per capita rate for tax totals per state. (tidycensus
    undocumented counts)?

-   Do states with favorable policies toward undocumented immigrants
    have friendly tax policies like rates (how will you define favorable
    policies)?

-   What are the other taxes undocumented immigrants pay and why do
    Californian undocumented immigrants pay almost 40% of nationwide
    other taxes.
