---
title: "Undocumented immigrants contribute $100 billion through taxes"
author: "Diego Perdomo for Futuro Media Group"
output: html_document
date: "2025-06-07"
editor_options: 
  markdown: 
    wrap: 72
---

In May, federal courts permitted immigration agencies to access
confidential, personal information of up to seven million undocumented
immigrant taxpayers. This is the first instance where the Internal
Revenue Service has shared confidential information for the purposes of
immigration investigation, according to the National Immigration Forum.

Required by law to file taxes regardless of citizenship, America's
undocumented immigrants, estimated at 11 million in 2022, are
responsible for nearly \$100 billion of contributions to federal, state
and local governments in 2022, according to a July report from Institute
on Taxation and Economic Policy.

<iframe title="NY undocumented pay highest rate to local, state taxes" aria-label="Map" id="datawrapper-chart-ADPu9" src="https://datawrapper.dwcdn.net/ADPu9/3/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="570" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r,i=0;r=e[i];i++)if(r.contentWindow===a.source){var d=a.data["datawrapper-height"][t]+"px";r.style.height=d}}}))}();
</script>

Since 1996, the IRS has permitted people with noncitizen status to file
their taxes through the Individual Taxpayer Identification Number. The
identification number, which permitted undocumented immigrants to file
their taxes and receive select tax credits, fell under scrutiny in
recent federal court hearings and prompted the landmark collaboration
between the Department of Homeland Security and Department of Treasury.

With new questions of undocumented immigrants' relationship to taxes and
rising confidentiality concerns, I think it would be timely and relevant
to Futuro’s audience to investigate:

-   Where the \$96.7 billion in contributions from undocumented
    immigrants goes to.

    -   Data visualizations like charts and maps to describe state-level
        rates for each tax revenue type (income, federal, sales, etc.)

-   How undocumented taxpayer contributions differ in makeup from
    citizen taxpayers

    -   The portfolio of taxes undocumented immigrants contribute
        compared to the U.S. total of tax revenue

Given the missing data on the exact counts of undocumented immigrants,
it would be interesting to examine:

-   How other tax policy think tanks measure immigrant contributions

    -   Particularly the methodology of these studies and what
        assumptions they make from the holes in their data.

-   Other instances of immigrant contribution to federal, state and
    local revenue

    -   Commerce and census data pertaining to income and nativity

Through this research, I intend to add new analysis to data that has
been discussed among spaces that report on immigrants and taxes for
almost 11 months and prepare information for followup articles as the
story develops.

# Load software libraries

```{r setup, include=FALSE}
#install.packages("tidyverse")
#install.packages("janitor")
#install.packages("readxl")
#install.packages("rvest")
library(tidyverse)
library(janitor)
library(readxl)
library(rvest)
library(DT)
```

# Scraping html tables from website

From \$96.7 billion, a third is spent on state and local taxes

<https://itep.org/undocumented-immigrants-taxes-2024/> APPENDIX TABLE 1.

State and Local Tax Contributions by the Undocumented Immigrant
Population in 2022

```{r}
#State local tax contribs from undoc immigrants in 2022
# Read the HTML file from the URL
url <- "https://itep.org/undocumented-immigrants-taxes-2024/"
html <- read_html(url)

# Extract the relevant information using XPath
# I'll show you how to do this.
data <- html %>%
  html_nodes(xpath = "/html/body/div/div/div[2]/div/div[1]/div[1]/div[4]/table") %>%
  html_nodes("td") %>%
  html_text() %>%
  matrix(ncol = 7, byrow = TRUE) %>%
  as.data.frame(stringsAsFactors = FALSE)

# Assign column names

new_headers <- as.character(data[1, ])
data_clean <- data[-1, ] # Remove the first row
colnames(data_clean) <- new_headers
```

<https://itep.org/undocumented-immigrants-taxes-2024/> APPENDIX TABLE 5.

```{r}
#Undocumented immigrants population and income data, 2022
undoc_immigrant_income_pop <- html |> 
  html_nodes(xpath = "/html/body/div[1]/div/div[2]/div/div[1]/div[1]/div[8]/table") |> 
  html_nodes("td") |> 
  html_text() |> 
  matrix(ncol = 3, byrow = TRUE) |> 
  as.data.frame(stringsAsFactors = FALSE)

###Make top consistent
undoc_immigrant_income_pop_clean <- undoc_immigrant_income_pop 

colnames(undoc_immigrant_income_pop_clean) <-     undoc_immigrant_income_pop[1,]

undoc_immigrant_income_pop_clean <- undoc_immigrant_income_pop_clean[-1,]
rownames(undoc_immigrant_income_pop_clean) <- 1:nrow(undoc_immigrant_income_pop_clean)


```

# Cleaning

```{r}
#Cleaning state local tax contribs from undoc immigrants in 2022
data <- data_clean |> 
  clean_names() |> 
  #remove dollar signs
  mutate(
    sales_and_excise_taxes=as.numeric(parse_number(sales_and_excise_taxes)),
    property_taxes=as.numeric(parse_number(property_taxes)),
    personal_and_business_income_taxes=as.numeric(parse_number(personal_and_business_income_taxes)),
    other_taxes=as.numeric(parse_number(other_taxes)),
    total_state_and_local_taxes=as.numeric(parse_number(total_state_and_local_taxes)),
     effective_tax_rate=as.numeric(parse_number(effective_tax_rate)),
    )

write.csv(data, "../data/state_local_tax_undoc_2022.csv")

#Remember to credit the source in your articles: Institute on Taxation and Economic Policy
```

```{r}
#Cleaning undocumented immigrant population and income data
undoc_immigrant_income_pop <- undoc_immigrant_income_pop_clean |> 
  clean_names() |> 
  #remove dollar signs
  mutate(
    population=as.numeric(parse_number(population)),
    aggregate_income=as.numeric(parse_number(aggregate_income))
    )

write_csv(undoc_immigrant_income_pop, "../data/aggregate_income_undoc_2022.csv")

#Remember to credit the source in your articles: Institute on Taxation and Economic Policy
```

```{r}
# Load your data

# YOUR_DATA <- read.csv("../data/XXXXX.csv") |> 
#     clean_names()

# data <- read_excel("../data/2024-12-Revenue.xlsx", skip=2) |> 
#     clean_names()

```

Data details:

# Explore the data types

```{r}

#glimpse(YOUR_DATA)

glimpse(data)

```

# Code for later Fix dates

```{r}
# data <- data |> 
#   mutate(date = lubridate::mdy(contribution_date))
```

# Data exploration: taxes paid

## State and Local Taxes

```{r echo=F}
##Decending Totals
#Total State and Local 
data |> 
  select(state, total_state_and_local_taxes) |> 
  arrange(desc(total_state_and_local_taxes))
```

```{r}
#Total Sales Tax 
data |> 
  select(state, sales_and_excise_taxes) |> 
  arrange(desc(sales_and_excise_taxes))

#Total Property 
data |> 
  select(state, property_taxes) |> 
  arrange(desc(property_taxes))
```

```{r hide=T,}
#Per capita
undoc_immigrant_taxes_income <- inner_join(data, undoc_immigrant_income_pop, by = "state")

#Sales - likely dependent on sales tax rates and population
undoc_immigrant_taxes_income <- undoc_immigrant_taxes_income |> 
  mutate(sales_capita = sales_and_excise_taxes/population,
         property_capita = property_taxes/population,
         personal_and_business_income_capita = personal_and_business_income_taxes/population,
         other_capita = other_taxes/population,
         total_capita = total_state_and_local_taxes/population) 

dt_undoc_tax_income <- undoc_immigrant_taxes_income |> 
  select(state,personal_and_business_income_capita,sales_capita, property_capita, total_capita) |> 
  rename(State="state",
         `Individual and Business Income Rate` = "personal_and_business_income_capita",
         `Sales Rate` = "sales_capita",
         `Property Rate` = "property_capita",
         `Total Rate` = "total_capita")

#Export and do visual analysis on datawrapper to visualize trends
write.csv(undoc_immigrant_taxes_income, "../data/taxes_capita_undoc_2022.csv")

datatable(dt_undoc_tax_income, caption = "Contributions to local and state taxes from undocumented immigrants", options = list(pageLength = 20)) |> 
  formatCurrency(c('Individual and Business Income Rate','Sales Rate','Property Rate','Total Rate'),currency = "$", interval = 3, mark = ",")
```

```{r hide=T}
# Other Taxes by State
data |> 
  select(state, other_taxes) |> 
  arrange(desc(other_taxes))


```

```{r}
#data |> 
```

# Questions

Write down questions you want answered:

-   What is the per capita rate for tax totals per state. (tidycensus
    undocumented counts)?

-   Do states with favorable policies toward undocumented immigrants
    have friendly tax policies like rates (how will you define favorable
    policies)?

-   What are the other taxes undocumented immigrants pay and why do
    Californian undocumented immigrants pay almost 40% of nationwide
    other taxes.

# Data Details

You will want to read this article carefully and you will probably want
to scrape more data from it.
<https://itep.org/undocumented-immigrants-taxes-2024/>

A very good article about the issue
<https://unidosus.org/blog/2025/03/24/when-dhs-requests-itin-tax-data-why-every-american-should-be-concerned/>

Background on ITIN
<https://www.irs.gov/tin/itin/individual-taxpayer-identification-number-itin>
If you're a resident alien, nonresident alien or their spouse or
dependent, you can apply for an ITIN regardless of immigration status.

Resident alien or nonresident alien

If you're not a U.S. citizen, your tax status can be either:

Resident alien – If you were present in the U.S. for more than 183
days\* (substantial presence test) or you’re a lawful permanent resident
of the U.S. (green card test) Nonresident alien – If you don’t meet
either the green card or substantial presence test for resident alien
status. \*Special day-counting rules apply to certain foreign
government-related individuals, teachers, trainees and students who are
temporarily present in the U.S. under certain U.S. visas.

Determining your tax residency status

U.S. Tax Guide for Aliens, Publication 519
